\chapter*{Table of Glossary}

\begin{longtable}{|p{3cm}|p{8cm}|p{3cm}|} 
\hline
\textbf{Term} & \textbf{Definitions} & \textbf{Abbreviation} \\
\hline
Adaptive Local Prototype & Prototype vectors that dynamically adjust to input variations. & ALP \\ 
\hline
Adaptive Local Prototype Network & Network that adapts local prototypes for improved segmentation. & ALPNet \\ 
\hline
Artificial Intelligence & Field encompassing machine learning, computer vision, and natural language processing. & AI \\
\hline
Class-Agnostic Network & Network that operates without requiring class-specific information. & CANet \\
\hline
Combined Healthy Abdominal Organ Segmentation & Dataset for abdominal organ segmentation in medical images. & CHAOS \\ 
\hline
Computed Tomography Scan & Medical imaging method for detailed internal visualization. & CT Scan \\
\hline
Connected Component Analysis & Identifies connected regions in binary images. & CCA \\ 
\hline
Convolutional Neural Networks & Deep learning models suited to grid-like data such as images. & CNNs \\
\hline
Dice Similarity Coefficient & Measures overlap between two segmentation outputs. & DSC \\ 
\hline
DICOM Format & Standard for storing and transmitting medical imaging data. & DICOM \\ 
\hline
Few-shot Segmentation & Segmentation using only a few annotated examples. & FSS \\ 
\hline
Floating Point Operations Per Second & Metric for computational performance. & FLOPs \\ 
\hline
Generative Adversarial Networks & Comprise a generator and discriminator in adversarial training to synthesize data. & GANs \\ 
\hline
Graphics Processing Unit & Hardware for parallel processing, commonly used in deep learning. & GPU \\ 
\hline
Local-Global & Combines local and global features for enhanced representation. & LoGo \\
\hline
Magnetic Resonance Imaging & Imaging technique using magnetic fields and radio waves. & MRI \\ 
\hline
Model-Agnostic Meta-Learning & Meta-learning method applicable across different models. & MAML \\ 
\hline
Natural Language Processing & Field focused on computational understanding of human language. & NLP \\ 
\hline
Prototype Alignment Network & Aligns class prototypes to enhance few-shot learning. & PANet \\ 
\hline
Prototype Alignment Regularization & Regularizes prototype learning for improved generalization. & PAR \\ 
\hline
Prototypical Networks & Metric-based few-shot learning using class prototypes in embedding space. & PN \\ 
\hline
Rectified Linear Unit & Non-linear activation function used in deep networks. & ReLU \\ 
\hline
Region of Interest & Specific image area selected for focused analysis. & ROI \\
\hline
Segment Anything Model & Foundation model capable of segmenting arbitrary image regions without task-specific training. & SAM \\ 
\hline
Similarity Guidance Network & Leverages similarity cues to guide segmentation. & SG-One \\ 
\hline
Vision Transformers & Transformer-based models for visual data, treating images as tokenized patches. & ViTs \\ 
\hline
\end{longtable}